---
title: "Week_3"
author: "Brennan"
date: "7/11/2019"
output: 
  html_document: 
    keep_md: yes
---

### Chapter 4: Workflow Basics
#### 4.1: Coding Basics
- using R as a calculator 
```{r}
(pi*32)/12
```
- you can create objects 
```{r}
x <- 3 *4
```
```{r}
x 
```
- format creating objects take: my_object <- value
#### 4.2: What's in a name?
- object names must start with a letter and contain only letters, numbers, and _ and . 
- using tab can help you cycle through your variables
- in order to call/use a variable, you have to type it the exact same way it was entered
- my_rocks and my_Rocks will not yield the same result
#### 4.3: Calling Functions
- if you R shows you +, that means that R is expecting you to input more, usually you've forgotten a " or a )
#### 4.4: Practice (One of the assigned problem sets)
1. Why does this code not work?
my_variable <- 10
my_varıable
- The i character is not the same in both variables. In order to use a variable, you have to call it by typying it the exact same way it was entered
2. Tweak each of the following commands so they run correctly.
library(tidyverse)
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
fliter(mpg, cyl = 8)
filter(diamond, carat > 3)
- Here are the correct versions:
```{r}
library("tidyverse")
```
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```
- Nothing wrong with this piece of code

filter(mpg, cyl == 8) +
filter(diamond, carat > 3)

3. Press Alt + Shift + K. What happens? How can you get to the same place using the menus?
- I do not have an alt key, but when I push contrl + Shift + k it allows me to save and name the file. I can get to the same place by going to file and then clicking Save As.

### Chapter 5: Data Transformation
#### 5.1: Introduction
```{r}
library("tidyverse")
```
```{r}
library("nycflights13")
```
- types of variables  
    - int: integers
    - dbl: doubles or real numbers
    - chr: character vectors or strings
    - dttm: date+time
    - lgl: logical vector
    - fctr: factors, categorical values with fixed possible values
    - date: dates
- dplyr basics  
    - filter(): pick observations by their values
    - arrange(): reorder the rows
    - select(): pick variables by their name
    - mutate(): create new variables with functions of existing variables
    - summarize(): collapse many values into a single summary 
    - group_by(): changes scope of each function from entire dataset to group by group 

#### 5.2: Filter()
```{r}
jan1 <- filter(flights, month == 1, day == 1)
```
```{r}
jan1
```
- in order to save the modified data frame, assign it to a variable
- to filter correctly remember the comparison operators: >, <, >=, <=, ==, and !=
- make sure you use ==, not =
- = will result in an error
Logical operators
- filter() also uses Boolean operators such as &(and), |(or), !(not)
```{r}
filter(flights, month == 11 | month == 12)
```
- another way to write the above function uses x%in%y
```{r}
nov_dec <- filter(flights, month %in% c(11, 12))
```
- do not use && and || with filter()
Missing values
- if you wanna determine if a value is missing use: is.na()
- filter() only includes values that are true for the condition, leaving out false and NA
Practice (one of the assigned problem sets)
1.1. Had an arrival delay of two or more hours
```{r}
filter(flights, dep_delay >= 120 )
```
1.4. Departed in summer (July, August, and September)
```{r}
filter(flights, month == 8 | month == 7 | month == 9 )
```
1.5. Arrived more than two hours late, but didn’t leave late
```{r}
filter(flights, dep_delay <=0 & arr_delay > 120 )
```
1.7. Departed between midnight and 6am (inclusive)
```{r}
filter(flights, dep_time>=0 & dep_time<=600 )
```
2. Another useful dplyr filtering helper is between(). What does it do? Can you use it to simplify the code needed to answer the previous challenges?
```{r}
?between
```
```{r}
filter(flights, between(dep_time, 0, 600) )
```
- between() is a shortcut for a>=x>=b. 
3. How many flights have a missing dep_time? What other variables are missing? What might these rows represent? 
```{r}
filter(flights, is.na(dep_time) )
```
- 1000; dep_delay, arr_time, arr_delay, airtime; Cancelled flights
#### 5.3: Arrange()
- arrange(): shifts the order the rows are in; the more column names you include in the function, the more specific the arrangement will be
```{r}
arrange(flights, year, month, day)
```
- desc(): is used to reorder a column in descending order
```{r}
arrange(flights, desc(dep_delay))
```
Exercises (one of the assigned problem sets)
1. How could you use arrange() to sort all missing values to the start? (Hint: use is.na()).
```{r}
arrange(flights, desc(is.na(dep_delay)))
```
2. Sort flights to find the most delayed flights. Find the flights that left earliest.
Most Delayed
```{r}
arrange(flights, desc(dep_delay))
```
Earliest
```{r}
arrange(flights, dep_delay)
```
3. Sort flights to find the fastest flights.
```{r}
arrange(flights, hour, minute, desc(distance))
```
4. Which flights travelled the longest? Which travelled the shortest?
Longest
```{r}
arrange(flights, desc(distance))
```
Shortest
```{r}
arrange(flights, (distance))
```
#### 5.4: Select()
- allows you to select and view specific columns
```{r}
select(flights, year, month, day)
```
- functions that can be used with select():
    - starts_with("abc"): matches with names that start with abc
    - ends_with("abc"): matches with names that end with abc
    - contains("abc"): matches with names that contain abc
    - matches("(.)\\1"): selects variables that match a regular expression. This one matches any variables that contain repeated characters
    - num_range("y", 1:4): matches y1, y2, y3, y4
- rename(): can be used to rename functions
```{r}
rename(flights, tail_num = tailnum)
```
- shifting the order of columns: everything() function in conjunction with select()
```{r}
select(flights, time_hour, air_time, everything())
```
Exercises (an assigned problem set)  
1. Brainstorm as many ways as possible to select dep_time, dep_delay, arr_time, and arr_delay from flights.
- select(flights, dep_time, dep_delay, arr_time, and arr_delay)
- select(flights, dep_time, dep_delay, arr_time, and arr_delay, everything())
- select(flights, starts_with("dep"), starts_with("arr"))
- select(flights, ends_with("time"), ends_with("delay"))
2. What happens if you include the name of a variable multiple times in a select() call?
```{r}
select(flights, time_hour, time_hour)
```
- The column is only called once.
3. What does the one_of() function do? Why might it be helpful in conjunction with this vector?
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
```{r}
?one_of
```
- matches variable names in a character vector. Instead of typing out year, month, day, dep_delay, and arr_delay, you would just assign it to variable vars and use one_of().
4. Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default? 
```{r}
select(flights, contains("TIME"))
```
- I am suprised by the result because I thought the uppercased version of time would throw things off. I guess function by default recognizes time and TIME as the same thing. I do not know how to change the default. CONFUSED
#### 5.5: Mutate()
- mutate() adds new columns to the end of the dataset that are functions of existing ones
- example:
```{r}
flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
)
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
)
```
- once you've created a column you can use it to create other columns
```{r}
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```
- if you only want to keep the new variable(s) use transmute():
```{r}
transmute(flights,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```
- modular arithmetic (allows you to break up integers into pieces)  
    - %/% (integer division)  
    - %% (remainder)  
- Logs: log(), log2(), log10()
- offsets: lead() or lag() allow you to refer to leading or lagging values
- cummulative aggregates:
    -cumsum(), cumprod(), cummin(), cummax(), cummean()
- Ranking:
    - min_rank(): ranks smallest values to highest
    - min_rank(desc()): ranks highest to smallest
    - variants: row_number(), dense_rank(), percent_rank(), cume_dist(), ntile()
Exercises (an assigned problem set)
1. Currently dep_time and sched_dep_time are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.

transmute(flights,
  deptime = dep_time,
  scheddeptime = 
)

- I have no Idea how to do this one. I know the function but what would convert it to normal minutes? CONFUSED
2. Compare air_time with arr_time - dep_time. What do you expect to see? What do you see? What do you need to do to fix it?
```{r}
mutate(flights,
  airtime = arr_time-dep_time
)
```
- I expected air_time and arr_time-dep_time to be close to the same. They are not that close together. arr_time and dep_time need to be converted into actual minutes instead of times.
3. Compare dep_time, sched_dep_time, and dep_delay. How would you expect those three numbers to be related?
- For the most part, I expect sched_dep_time + dep_delay to equal dep_time. However, dep_time and sched_dep_time being times instead of minutes complicates things because minutes delayed may not translate well depending on how long the flight was delayed. 
4. Find the 10 most delayed flights using a ranking function. How do you want to handle ties? Carefully read the documentation for min_rank().
- NEED HELP. I don't know how to put flights into a vector format for min_rank
5. What does 1:3 + 1:10 return? Why?
```{r}
1:3 + 1:10
```
- 10 is not a multiple of 3 so the shorter object can't repeat itself to match the longer object's length
#### 5.6: Grouped summaries with summarize()
- collapses a data frame into a single row
```{r}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
```
- you can pair summarise() with group_by() in order to change the unit of analysis from entire data set to individual groups
```{r}
by_day <- group_by(flights, year, month, day)
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```
- Combining operations with %>%  
- original:  
  by_dest <- group_by(flights, dest)  
    delay <- summarise(by_dest,  
    count = n(),  
    dist = mean(distance, na.rm = TRUE),  
    delay = mean(arr_delay, na.rm = TRUE)  
  )  
  delay <- filter(delay, count > 20, dest != "HNL")  
  ggplot(data = delay, mapping = aes(x = dist, y = delay)) +  
    geom_point(aes(size = count), alpha = 1/3) +  
    geom_smooth(se = FALSE)  
- with %>%:   
delays <- flights %>%   
  group_by(dest) %>%   
  summarise(  
    count = n(),  
    dist = mean(distance, na.rm = TRUE),  
    delay = mean(arr_delay, na.rm = TRUE)  
  ) %>%   
  filter(count > 20, dest != "HNL")  
- %>% essentially acts as "then"; this code essentially reads group the data, then summarise, then filter  
Missing values
  - na.rm = TRUE removes all the missing values prior to computation  
  - another way to remove NA(using filter(!is.na())):  filter(!is.na(dep_delay), !is.na(arr_delay))
Useful Summary functions
- Measures of Location:
    - mean(): sum divided by length
    - median(): middle of the data
- Measures of spread
    - sd(): standard deviation 
    - IQR(): inner quartile range 
    - mad(): median absolute deviation 
- Measures of Rank
    - min(): minimum
    - max(): maximum
    - quantile ( , .25): will find value that is greater than 25% of values and less than the remaining 75%
- Measures of Position
    - first(): first
    - last(): last
    - nth( , 2): 2nd
- Counts
    - n(): counts the size of the current group
    - sum(!is.na()): count number of non-missing values
    - n_distinct(): counts number of unique values
- you can ungroup data wth the ungroup() function
Exercises (an assigned problem set)
5. Which carrier has the worst delays? Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about flights %>% group_by(carrier, dest) %>% summarise(n()))
```{r}
worst_delays <- group_by(flights,  carrier)
summarise(worst_delays, delay = mean(dep_delay, na.rm = TRUE))
```
```{r}
worst_delays <- group_by(flights,  dest)
summarise(worst_delays, delay = mean(dep_delay, na.rm = TRUE))
```
```{r}
worst_delays <- group_by(flights,  carrier, dest)
summarise(worst_delays, delay = mean(dep_delay, na.rm = TRUE))
```
- No you can't disentangle the effects of bad airports vs. bad carriers. While looking at the data, it is hard to tell whether the delays are the result of the airport or the carrier. You have to look at them together because the flight's performance is a combination of the carrier's quality and aiport quality. 
#### 5.7: Grouped Mutates and Filters
- finding the worst members of each grouping
```{r}
flights_sml %>% 
  group_by(year, month, day) %>%
  filter(rank(desc(arr_delay)) < 10)
```
- finding all groups bigger than a threshold
```{r}
popular_dests <- flights %>% 
  group_by(dest) %>% 
  filter(n() > 365)
popular_dests
```
- standardizing to compute per group metrics
```{r}
popular_dests %>% 
  filter(arr_delay > 0) %>% 
  mutate(prop_delay = arr_delay / sum(arr_delay)) %>% 
  select(year:day, dest, arr_delay, prop_delay)
```

